Synthesis of Linear Functional Programs

### 2021-07-23 article/article.tex
final :)
Adjustments pos-delivery

% TODO: Temos de ter o título assim? ficava mais fixe sem o número e sem o
% supervisor :P
\title{Synthesis of Linear Functional Programs}
\author{Rodrigo Mesquita - Nº 55902 \\
    Supervisor: Bernardo Toninho}
\date{July 2021}
\institute{NOVA School of Science and Technology}

@@

The \synname\ synthesizer combines linear types with recursion,
parametric polymorphism, recursive algebraic data types, and
refinement types. The synthesizer is built on top of a system of proof-search for
linear logic. Proof search relates to program synthesis via the Curry-Howard
correspondence~\cite{curry:34,howard:80,DBLP:journals/cacm/Wadler15}, which states that \emph{propositions in a logic} are
\emph{types}, and \emph{proofs} of those propositions are well-typed
\emph{programs} -- finding a proof of a proposition is finding a program with
that type.

@@

\begin{mathpar}
    \infer*[right=($\forall L$)]
    {
        \Theta/\Theta'; \Rho; \Gamma; \Delta/\Delta'; \tau' \Downarrow\ \vdash C
        \\
        \forall \overline{\alpha}.\ \tau \sqsubseteq_E \tau'
        \\
        \textrm{ftv}_E(\tau') \cap \{ ?\alpha\ \vert\ (?\alpha \mapsto \tau_x) \in \Theta'\} = \emptyset
    }
    {\Theta/\Theta'; \Rho; \Gamma; \Delta/\Delta'; \forall \overline{\alpha}.\ \tau \Downarrow\ \vdash C}
    \and
    \infer*[right=($?L$)]
    {\textsc{unify}(?\alpha
    \mapsto C, \Theta)}
    {\Theta/\Theta, ?\alpha \mapsto C ; \Rho; \Gamma; \Delta/\Delta';
    x{:}?\alpha \Downarrow\ \vdash x : C}
    \and
    \infer*[right=($\Downarrow L?$)]
    {\textsc{unify}(?\alpha
    \mapsto A, \Theta)}
    {\Theta/\Theta, ?\alpha \mapsto A ; \Rho; \Gamma; \Delta/\Delta';
    x{:}A \Downarrow\ \vdash x : ?\alpha}
\end{mathpar}

@@

@@

\noindent \emph{Example.} To showcase the expressiveness of our synthesizer, we present as
examples the input and output program for \emph{reverse}.
Note how we can guide the synthesis using the assertion mechanism to 
supply an example:

@@

    {\Gamma, A; \Delta/\Delta' ; A \Downarrow\ \vdash C}
    {\Gamma, A; \Delta/\Delta';\cdot \Uparrow\ \vdash C}

@@

    {\Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ C_n \ M : T \Downarrow}
        \Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        \Gamma ; \Delta/ \Delta'_2 ; \Omega, y_2{:}X_2 \Uparrow\ \vdash M_2 : C
        y_2 \notin \Delta'_2
        \Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        \Delta'_1 = \Delta'_2 = \dots = \Delta'_n
    {\Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}T \Uparrow\

@@

    {\Rho; \Gamma; \Delta/\Delta'; x{:}T \Downarrow\ \vdash x : T}

@@

    \infer*[right=($\Downarrow L?$)]

@@

insert :: forall a . (a -o (List a -o List a));
insert b c = case c of
    Nil -> Cons (b, Nil)
  | Cons g ->
      let h*i = g in
        Cons (h, insert b i);

concat :: forall a . (List (List a) -o List a);
concat b = case b of
    Nil -> Nil
  | Cons c ->
      let d*e = c in
        case d of
            Nil -> concat e
          | Cons k ->
              let l*m = k in
                Cons (l, concat (Cons (m, e)));
\end{minted}

>>>>

% TODO: Temos de ter o título assim? ficava mais fixe sem o número e sem o
% supervisor :P
\title{Synthesis of Linear Functional Programs}
\author{Rodrigo Mesquita \and Bernardo Toninho}
\date{July 2021}
\institute{NOVA School of Science and Technology}

@@

The \synname\ synthesizer combines linear types with recursion,
parametric polymorphism, recursive algebraic data types, and
refinement types. The synthesizer is built on top of a system of proof-search for
linear logic. Proof search relates to program synthesis via the Curry-Howard
correspondence~\cite{curry:34,howard:80,DBLP:journals/cacm/Wadler15}, which states that \emph{propositions in a logic} are
\emph{types}, and their \emph{proofs} are well-typed
\emph{programs} -- finding a proof of a proposition is finding a program with
that type.

@@

\begin{mathpar}
    \infer*[right=($\forall L$)]
    {
        \Theta/\Theta'; \Rho; \Gamma; \Delta/\Delta'; \tau' \Downarrow\ \vdash C
        \\
        \forall \overline{\alpha}.\ \tau \sqsubseteq_E \tau'
        \\
        \textrm{ftv}_E(\tau') \cap \{ ?\alpha\ \vert\ (?\alpha \mapsto \tau_x) \in \Theta'\} = \emptyset
    }
    {\Theta/\Theta'; \Rho; \Gamma; \Delta/\Delta'; \forall \overline{\alpha}.\ \tau \Downarrow\ \vdash C}
    \and
    \infer*[right=($?L$)]
    {\textsc{unify}(?\alpha
    \mapsto C, \Theta)}
    {\Theta/\Theta, ?\alpha \mapsto C ; \Rho; \Gamma; \Delta/\Delta';
    x{:}?\alpha \Downarrow\ \vdash x : C}
    \and
    \infer*[right=($\Downarrow ?L$)]
    {\textsc{unify}(?\alpha
    \mapsto A, \Theta)}
    {\Theta/\Theta, ?\alpha \mapsto A ; \Rho; \Gamma; \Delta/\Delta';
    x{:}A \Downarrow\ \vdash x : ?\alpha}
\end{mathpar}

@@

\mypara{Example} To showcase the expressiveness of our synthesizer, we present as
examples the input and output program for \emph{reverse}.
Note how we can guide the synthesis using the assertion mechanism to 
supply an example:

@@

@@

@@

@@

@@

    \infer*[right=($\Downarrow ?L$)]

@@

insert :: forall a . (a -o (List a -o List a));
insert b c = case c of
    Nil -> Cons (b, Nil)
  | Cons h*i -> Cons (h, insert b i);

concat :: forall a . (List (List a) -o List a);
concat b = case b of
    Nil -> Nil
  | Cons d*e ->
        case d of
            Nil -> concat e
          | Cons k ->
              let l*m = k in
                Cons (l, concat (Cons (m, e)));
\end{minted}
